# The WCGS {#sec-WCGS}

## Setup: Packages Used Here

```{r}
#| message: false

knitr::opts_chunk$set(comment = NA)

library(broom)
library(GGally)
library(ggridges)
library(janitor)
library(kableExtra)
library(patchwork)
library(tidyverse)

theme_set(theme_bw())
```

We will also use the `favstats` function from the `mosaic` package, even though I won't load `mosaic` here.

## The Western Collaborative Group Study (`wcgs`)

@Vittinghoff explore data from the Western Collaborative Group Study (WCGS) in some detail^[For more on the WCGS, you might look at http://www.epi.umn.edu/cvdepi/study-synopsis/western-collaborative-group-study/]. We'll touch lightly on some key issues in this Chapter.

> The Western Collaborative Group Study (WCGS) was designed to test the hypothesis that the so-called Type A behavior pattern (TABP) - "characterized particularly by excessive drive, aggressiveness, and ambition, frequently in association with a relatively greater preoccupation with competitive activity, vocational deadlines, and similar pressures" - is a cause of coronary heart disease (CHD). Two additional goals, developed later in the study, were (1) to investigate the comparability of formulas developed in WCGS and in the Framingham Study (FS) for prediction of CHD risk, and (2) to determine how addition of TABP to an existing multivariate prediction formula affects ability to select subjects for intervention programs.

The study enrolled over 3,000 men ages 39-59 who were employed in San Francisco or Los Angeles, during 1960 and 1961.

In the code chunk below, after importing the data and creating a tibble with `read_csv`, I used `mutate(across(where(is.character), as_factor)` to convert all variables containing character data into factors.

```{r}
#| message: false

wcgs <- read_csv("data/wcgs.csv") |>
    mutate(across(where(is.character), as_factor))

wcgs
```

Here, we have `r dim(wcgs)[1]` rows (subjects) and `r dim(wcgs)[2]` columns (variables). 

### Structure of `wcgs`

We can specify the (sometimes terrible) variable names, through the `names` function, or we can add other elements of the structure, so that we can identify items of particular interest.

```{r}
str(wcgs)
```

### Codebook for `wcgs`

This table was lovingly hand-crafted, and involved a lot of typing. We'll look for better ways in 432.

Name    | Stored As  | Type         | Details (units, levels, etc.)
-------:|:----------:|:------------:|------------------------------------
`id`      | integer    | (nominal)  | ID #, nominal and uninteresting
`age`     | integer    | quantitative | age, in years - no decimal places
`agec`    | factor (5) | (ordinal) | age: 35-40, 41-45, 46-50, 51-55, 56-60
`height`  | integer    | quantitative | height, in inches
`weight`  | integer    | quantitative | weight, in pounds
`lnwght`  | number     | quantitative | natural logarithm of weight
`wghtcat` | factor (4) | (ordinal) | wt: < 140, 140-170, 170-200, > 200
`bmi`     | number     | quantitative | body-mass index: <br /> 703 * weight in lb / (height in in)^2^
`sbp`     | integer    | quantitative | systolic blood pressure, in mm Hg
`lnsbp`   | number     | quantitative | natural logarithm of sbp
`dbp`     | integer    | quantitative | diastolic blood pressure, mm Hg
`chol`    | integer    | quantitative | total cholesterol, mg/dL
`behpat`  | factor (4) | (nominal) | behavioral pattern: A1, A2, B3 or B4
`dibpat`  | factor (2) | (binary)  | behavioral pattern: A or B
`smoke`   | factor (2) | (binary)  | cigarette smoker: Yes or No
`ncigs`   | integer | quantitative | number of cigarettes smoked per day
`arcus`   | integer | (nominal) | arcus senilis present (1) or absent (0)
`chd69`   | factor (2) | (binary) | CHD event: Yes or No
`typchd69`| integer | (4 levels) | event: 0 = no CHD, 1 = MI or SD, <br /> 2 = silent MI, 3 = angina
`time169` | integer | quantitative | follow-up time in days
`t1`      | number  | quantitative | heavy-tailed (random draws)
`uni`     | number  | quantitative | light-tailed (random draws)

### Quick Summary

```{r}
summary(wcgs)
```

For a more detailed description, we might consider `Hmisc::describe`, `psych::describe`, `mosaic::inspect`, etc., as we've done (for instance) in @sec-pen2 and @sec-summ_quant.

## Are the SBPs Normally Distributed?

Consider the question of whether the distribution of the systolic blood pressure results is well-approximated by the Normal, where we'll make use of tools based on our discussion in @sec-normal.

```{r}
res <- mosaic::favstats(~ sbp, data = wcgs)
bin_w <- 5 # specify binwidth

ggplot(wcgs, aes(x = sbp)) +
    geom_histogram(binwidth = bin_w, 
                   fill = "orchid", 
                   col = "blue") +
    stat_function(
        fun = function(x) dnorm(x, mean = res$mean, 
                                sd = res$sd) * 
            res$n * bin_w,
        col = "navy") +
    labs(title = "Systolic BP for `wcgs` subjects",
     x = "Systolic BP (mm Hg)", y = "",
     caption = "Superimposed Normal model")
```

Since the data contain both `sbp` and `lnsbp` (its natural logarithm), let's compare them. Note that in preparing the graph, we'll need to change the location for the text annotation.

```{r}
res <- mosaic::favstats(~ lnsbp, data = wcgs)
bin_w <- 0.05 # specify binwidth

ggplot(wcgs, aes(x = lnsbp)) +
    geom_histogram(binwidth = bin_w, 
                   fill = "orange", 
                   col = "blue") +
    stat_function(
        fun = function(x) dnorm(x, mean = res$mean, 
                                sd = res$sd) * 
            res$n * bin_w,
        col = "navy") +
    labs(title = "ln(Systolic BP) for `wcgs` subjects",
     x = "ln(Systolic BP)", y = "",
     caption = "Superimposed Normal model")
```

We can also look at Normal Q-Q plots, for instance...

```{r}
p1 <- ggplot(wcgs, aes(sample = sbp)) +
    geom_qq(color = "orchid") + 
    geom_qq_line(color = "red") +
    labs(y = "Ordered SBP", title = "sbp in wcgs")

p2 <- ggplot(wcgs, aes(sample = lnsbp)) +
    geom_qq(color = "orange") + 
    geom_qq_line(color = "red") +
    labs(y = "Ordered ln(SBP)", title = "ln(sbp) in wcgs")

## next step requires library(patchwork)

p1 + p2 + 
    plot_annotation(title = "Normal Q-Q plots of SBP and ln(SBP) in wcgs")
```

There's at best a small improvement from `sbp` to `lnsbp` in terms of approximation by a Normal distribution.

## Identifying and Describing SBP outliers

It looks like there's an outlier (or a series of them) in the SBP data.

```{r}
#| fig-height: 4

ggplot(wcgs, aes(x = "", y = sbp)) +
    geom_violin() +
    geom_boxplot(width = 0.3, fill = "royalblue", 
                 outlier.color = "royalblue") +
    labs(title = "Boxplot with Violin of SBP in `wcgs` data",
         y = "Systolic Blood Pressure (mm Hg)", 
         x = "") +
    coord_flip() 
```

```{r}
mosaic::favstats(wcgs$sbp)
Hmisc::describe(wcgs$sbp)
```

The maximum value here is `r max(wcgs$sbp)`, and is clearly the most extreme value in the data set. One way to gauge this is to describe that observation's **Z score**, the number of standard deviations away from the mean that the observation falls. Here, the maximum value, `r max(wcgs$sbp)` is `r round_half_up((max(wcgs$sbp)-mean(wcgs$sbp))/sd(wcgs$sbp), 2)` standard deviations above the mean, and thus has a Z score of `r round_half_up((max(wcgs$sbp)-mean(wcgs$sbp))/sd(wcgs$sbp), 1)`.

A negative Z score would indicate a point below the mean, while a positive Z score indicates, as we've seen, a point above the mean. The minimum systolic blood pressure, `r min(wcgs$sbp)` is `r round_half_up((mean(wcgs$sbp)-min(wcgs$sbp))/sd(wcgs$sbp), 2)` standard deviations *below* the mean, so it has a Z score of `r round_half_up((min(wcgs$sbp)-mean(wcgs$sbp))/sd(wcgs$sbp), 1)`. 

Recall that the Empirical Rule (described in @sec-normal) suggests that if a variable follows a Normal distribution, it would have approximately 95% of its observations falling inside a Z score of (-2, 2), and 99.74% falling inside a Z score range of (-3, 3). Do the systolic blood pressures appear Normally distributed?

## Does Weight Category Relate to SBP?

The data are collected into four groups based on the subject's weight (in pounds).

```{r}
ggplot(wcgs, aes(x = wghtcat, y = sbp)) +
    geom_violin() +
    geom_boxplot(aes(fill = wghtcat), width = 0.3, notch = TRUE) +
    scale_fill_viridis_d() +
    guides(fill = "none") + 
    labs(title = "Boxplot of Systolic BP by Weight Category in WCGS", 
         x = "Weight Category", y = "Systolic Blood Pressure")
```

## Re-Leveling a Factor

Well, that's not so good. We really want those weight categories (the *levels*) to be ordered more sensibly. 

```{r}
wcgs |> tabyl(wghtcat)
```

Like all *factor* variables in R, the categories are specified as levels. We want to change the order of the levels in a new version of this factor variable so they make sense. There are multiple ways to do this, but I prefer the `fct_relevel` function from the `forcats` package (part of the tidyverse.) Which order is more appropriate?

I'll add a new variable to the `wcgs` data called `weight_f` that relevels the `wghtcat` data.

```{r}
wcgs <- wcgs |>
    mutate(weight_f = fct_relevel(wghtcat, "< 140", "140-170", "170-200", "> 200"))

wcgs |> tabyl(weight_f)
```

For more on the `forcats` package, check out @R4DS, especially its [section on Factors](https://r4ds.hadley.nz/factors.html).

### SBP by Weight Category

```{r}
ggplot(wcgs, aes(x = weight_f, y = sbp, fill = weight_f)) +
    geom_boxplot(notch = TRUE) +
    scale_fill_viridis_d() +
    guides(fill = "none") +
    labs(title = "Systolic Blood Pressure by Reordered Weight Category in WCGS", 
         x = "Weight Category", y = "Systolic Blood Pressure")
```

We might see some details well with a **ridgeline plot**, too.

```{r}
ggplot(wcgs, aes(x = sbp, y = weight_f, fill = weight_f, height = ..density..)) +
    ggridges::geom_density_ridges(scale = 2) +
    scale_fill_viridis_d() +
    guides(fill = "none") +
    labs(title = "SBP by Weight Category (wcgs)",
         x = "Systolic Blood Pressure",
         y = "Weight Category") 
```

As the plots suggest, patients in the heavier groups generally had higher systolic blood pressures.

```{r}
mosaic::favstats(sbp ~ weight_f, data = wcgs)
```

## Are Weight and SBP Linked?

Let's build a scatter plot of SBP (Outcome) by Weight (Predictor), rather than breaking down into categories.

```{r}
#| message: false

ggplot(wcgs, aes(x = weight, y = sbp)) +
    geom_point(size=3, shape=1, color="forestgreen") + ## default size = 2
    geom_smooth(method = "lm", se = FALSE, col = "red", formula = y ~ x) +
    geom_smooth(method = "loess", col = "blue", formula = y ~ x) +
    ggtitle("SBP vs. Weight in 3,154 WCGS Subjects") 
```

- The mass of the data is hidden from us - showing `r length(wcgs$sbp)` points in one plot can produce little more than a blur where there are lots of points on top of each other.
- Here the least squares regression line (in red), and loess scatterplot smoother, (in blue) can help.

The relationship between systolic blood pressure and weight appears to be very close to linear, but of course there is considerable scatter around that generally linear relationship. It turns out that the Pearson correlation of these two variables is `r signif(cor(wcgs$sbp, wcgs$weight),3)`.

## SBP and Weight by Arcus Senilis groups?

An issue of interest to us will be to assess whether the SBP-Weight relationship we see above is similar among subjects who have been diagnosed with arcus senilis and those who have not. 

> Arcus senilis is an old age syndrome where there is a white, grey, or blue opaque ring in the corneal margin (peripheral corneal opacity), or white ring in front of the periphery of the iris. It is present at birth but then fades; however, it is quite commonly present in the elderly. It can also appear earlier in life as a result of hypercholesterolemia.
> 
> Wikipedia article on Arcus Senilis, retrieved 2017-08-15

Let's start with a quick look at the `arcus` data.

```{r}
wcgs |> tabyl(arcus)
```

We have `r sum(is.na(wcgs$arcus))` missing values, so we probably want to do something about that before plotting the data, and we may also want to create a factor variable with more meaningful labels than 1 (which means yes, arcus senilis is present) and 0 (which means no, it isn't.)

```{r}
wcgs <- wcgs |>
    mutate(arcus_f = fct_recode(factor(arcus),
                                "Arcus senilis" = "1",
                                "No arcus senilis" = "0"),
           arcus_f = fct_relevel(arcus_f, "Arcus senilis"))

wcgs |> tabyl(arcus_f, arcus)
```

Let's build a version of the `wcgs` data that eliminates all missing data in the variables of immediate interest, and then plot the SBP-weight relationship in groups of patients with and without arcus senilis.

```{r}
#| message: false

wcgs_temp <- wcgs |>
    filter(complete.cases(arcus_f, sbp, weight)) 

ggplot(wcgs_temp, aes(x = weight, y = sbp, group = arcus_f)) +
    geom_point(shape = 1) + 
    geom_smooth(method = "lm", col = "red", formula = y ~ x) +
    geom_smooth(method = "loess", se = FALSE, col = "blue", formula = y ~ x) +
    labs(title = "SBP vs. Weight by Arcus Senilis status",
         caption = "3,152 WCGS with known arcus senilis status") + 
    facet_wrap(~ arcus_f) 
```

## Linear Model for SBP-Weight Relationship: subjects without Arcus Senilis

```{r}
model.noarcus <- 
    lm(sbp ~ weight, data = filter(wcgs, arcus == 0))

tidy(model.noarcus) |> 
    kbl(digits = 2) |>
    kable_styling(full_width = FALSE)

glance(model.noarcus) |> 
    select(r.squared:p.value, AIC) |> 
    kbl(digits = c(3, 3, 1, 1, 3, 0)) |>
    kable_styling(full_width = FALSE)

summary(model.noarcus)
```

The linear model for the `r table(wcgs$arcus)["0"]` patients without Arcus Senilis has R-squared = `r 100*signif(summary(model.noarcus)$r.square,3)`\%.

- The regression equation is 95.92 - 0.19 weight, for those patients without Arcus Senilis.

## Linear Model for SBP-Weight Relationship: subjects with Arcus Senilis

```{r}
model.witharcus <- 
    lm(sbp ~ weight, data = filter(wcgs, arcus == 1))

tidy(model.witharcus) |> 
    kbl(digits = 2) |>
    kable_styling(full_width = FALSE)

glance(model.witharcus) |> 
    select(r.squared:p.value, AIC) |> 
    kbl(digits = c(3, 3, 1, 1, 3, 0)) |>
    kable_styling(full_width = FALSE)

summary(model.witharcus)
```

The linear model for the `r table(wcgs$arcus)["1"]` patients with Arcus Senilis has R-squared = `r 100*signif(summary(model.witharcus)$r.square,3)`\%.

- The regression equation is 101.88 - 0.163 weight, for those patients with Arcus Senilis.

## Including Arcus Status in the model

```{r}
model3 <- lm(sbp ~ weight * arcus, data = filter(wcgs, !is.na(arcus)))

tidy(model3) |> 
    kbl(digits = 2) |>
    kable_styling(full_width = FALSE)

glance(model3) |> 
    select(r.squared:p.value, AIC) |> 
    kbl(digits = c(3, 3, 1, 1, 3, 0)) |>
    kable_styling(full_width = FALSE)

summary(model3)
```

The actual regression equation in this setting includes both weight, and an indicator variable (1 = yes, 0 = no) for arcus senilis status, and the product term combining weight and that 1/0 indicator. In 432, we'll spend substantial time and energy discussing these product terms, but we'll not do much of that in 431.

- Note the use of the product term `weight*arcus` in the setup of the model to allow both the slope of weight and the intercept term in the model to change depending on arcus senilis status.
    + For a patient who has arcus, the regression equation is SBP = 95.92 + 0.19 weight + 5.96 (1) - 0.028 weight (1) = 101.88 + 0.162 weight.
    + For a patient without arcus senilis, the regression equation is SBP = 95.92 + 0.19 weight + 5.96 (0) - 0.028 weight (0) = 95.92 + 0.19 weight. 

The linear model including the interaction of weight and arcus to predict sbp for the `r sum(!is.na(wcgs$arcus))` patients with known Arcus Senilis status has R-squared = `r 100*signif(summary(model3)$r.square,3)`\%. Again, we'll discuss interaction more substantially in 432.


## Predictions from these Linear Models

What is our predicted SBP for a subject weighing 175 pounds?
    
How does that change if our subject weighs 200 pounds?
    
Recall that

- *Without* Arcus Senilis, linear model for SBP = `r round_half_up(coef(model.noarcus)[1],1)` + `r round_half_up(coef(model.noarcus)[2],2)` x weight
- *With* Arcus Senilis, linear model for SBP = `r round_half_up(coef(model.witharcus)[1],1)` + `r round_half_up(coef(model.witharcus)[2],2)` x weight

So the predictions for a 175 pound subject are:

- `r round_half_up(coef(model.noarcus)[1],1)` + `r round_half_up(coef(model.noarcus)[2],2)` x 175 = `r round_half_up(coef(model.noarcus)[1] + coef(model.noarcus)[2]*175,0)` mm Hg without Arcus Senilis, and 

- `r round_half_up(coef(model.witharcus)[1],1)` + `r round_half_up(coef(model.witharcus)[2],2)` x 175 = `r round_half_up(coef(model.witharcus)[1] + coef(model.witharcus)[2]*175,0)` mm Hg with Arcus Senilis.

And thus, the predictions for a 200 pound subject are:

- `r round_half_up(coef(model.noarcus)[1],1)` + `r round_half_up(coef(model.noarcus)[2],2)` x 200 = `r round_half_up(coef(model.noarcus)[1] + coef(model.noarcus)[2]*200,1)` mm Hg without Arcus Senilis, and 

- `r round_half_up(coef(model.witharcus)[1],1)` + `r round_half_up(coef(model.witharcus)[2],2)` x 200 = `r round_half_up(coef(model.witharcus)[1] + coef(model.witharcus)[2]*200,1)` mm Hg with Arcus Senilis.

## Scatterplots with Facets Across a Categorical Variable

We can use facets in `ggplot2` to show scatterplots across the levels of a categorical variable, like `behpat`.

```{r, message = FALSE}
ggplot(wcgs, aes(x = weight, y = sbp, col = behpat)) +
    geom_point() +
    facet_wrap(~ behpat) +
    geom_smooth(method = "lm", se = FALSE, 
                formula = y ~ x, col = "black") +
    guides(color = "none") +
    theme(strip.text = element_text(face="bold", size=rel(1.25), color="white"),
          strip.background = element_rect(fill="royalblue")) +
    labs(title = "Scatterplots of SBP vs. Weight within Behavior Pattern")
```

## Scatterplot and Correlation Matrices

A **scatterplot matrix** can be very helpful in understanding relationships between multiple variables simultaneously. There are several ways to build such a thing, including the `pairs` function...

```{r}
#| fig-height: 6

pairs (~ sbp + age + weight + height, data=wcgs, 
       main="Simple Scatterplot Matrix")
```

### Displaying a Correlation Matrix

```{r}
wcgs |>
    select(sbp, age, weight, height) |>
    cor() |>
    kbl(digits = 3) |>
    kable_styling(full_width = FALSE)
```

### Using the `GGally` package

The `ggplot2` system doesn't have a built-in scatterplot system. There are some nice add-ins in the world, though. One option I sort of like is in the `GGally` package, which can produce both correlation matrices and scatterplot matrices.

The `ggpairs` function provides a density plot on each diagonal, Pearson correlations on the upper right and scatterplots on the lower left of the matrix.

```{r}
#| message: false

ggpairs(wcgs |> 
          select(sbp, age, weight, height),
        title = "Scatterplot Matrix via ggpairs")
```



