# Setting Up R {.unnumbered}

These Notes make extensive use of

-   the statistical [software language R](https://cran.case.edu/), and
-   the development environment [RStudio](https://www.rstudio.com/products/rstudio/),

both of which are free, and you'll need to install them on your machine. Instructions for doing so will be found on [the course website](https://thomaselove.github.io/431-2023/).

If you need a gentle introduction, or if you're just new to R and RStudio and need to learn about them, we encourage you to take a look at [https://moderndive.com/](http://moderndive.com/), which provides an introduction to statistical and data sciences via R at @ModernDive.

## Quarto

These notes were written using [Quarto](https://quarto.org/), which we'll learn to use in 431. Quarto, like R and RStudio, is free and open source.

Quarto is described as an *scientific and technical publishing system* for data science, which lets you (among many other things):

-   save and execute R code
-   generate high-quality reports that can be shared with an audience

The [Quarto Get Started page](https://quarto.org/docs/get-started/) which provides an overview and quick tour of what's possible with Quarto.

Another excellent resource to learn more about Quarto is the Communicate section (especially the [Quarto chapter]https://r4ds.hadley.nz/quarto.html)) of @R4DS.

## R Packages {.unnumbered}

At the start of each chapter that involves R code, I'll present a series of commands I run to set up R to use several packages (libraries) of functions that expand its capabilities, make a specific change to how I want R output to be displayed (that's the `comment = NA` piece) and sets the theme for most graphs to `theme_bw()`. A chunk of code like this will occur near the top of any Quarto work.

For example, this is the setup for one of our early chapters that loads four packages.

```{r setup, message=FALSE}
knitr::opts_chunk$set(comment = NA)

library(palmerpenguins)
library(janitor)
library(knitr)
library(tidyverse)

theme_set(theme_bw())
```

You only need to install a package once, but you need to reload it (using the `library()` function) every time you start a new session. I always load the package called `tidyverse` last, since doing so avoids some annoying problems.

## The `Love-boost.R` script {.unnumbered}

In October, when we start Part B of the course, we'll use some special R functions I've gathered for you in a script called `Love-boost`. I'll tell R about that code using the following command...

```{r}
source("data/Love-boost.R")
```

The `Love-boost.R` script includes four functions:

- `bootdif`
- `saifs.ci`
- `twobytwo`
- `retrodesign`

## Packages Used in these Notes {.unnumbered}

A list of all R packages we want you to install this semester (which includes some packages not included in these Notes) is maintained at [our course web site](https://thomaselove.github.io/431-2023/).

Package | Parts | Key functions in the Package
:-------: | :---: | :---: 
`arm` | C | --
`boot` | B | --
`broom` | A, B, C | `tidy`, `glance`, `augment` (part of `tidymodels`)
`car` | A, C |  `boxCox`, `powerTransform`
`Epi` | B | `twoby2`
`fivethirtyeight` | Appendix | source of data
`GGally` | A, C | `ggpairs`
`ggrepel` | C | --
`ggridges` | A, B | --
`ggstance` | A | --
`gt` | A | for presenting tables
`gtsummary` | A | `tbl_summary`
`Hmisc` | A, B, C | `describe` and others
`janitor` | A, B, C | `tabyl` and others
`kableExtra` | A | `kbl`, `kable_stylings`
`knitr` | A, B, C | `kable`
`lvplot` | A | `geom_lv`
`mice` | C | --
`modelsummary` | A, C | `modelsummary`
`mosaic` | A, B, C | `favstats`, `inspect`
`naniar` | A | `n_miss`, `miss_case_table`, `gg_miss_var`
`NHANES` | A | source of data
`palmerpenguins` | A | source of data
`patchwork` | A, B, C | for combining/annotating plots
`psych` | A, B | `describe`
`pwr` | B | --
`rms` | C | --
`sessioninfo` | Appendix | --
`simputation` | A | various imputation functions
`summarytools` | A | `descr`, `dfSummary`
`tidyverse` | A, B, C, Appendix | dozens of functions
`vcd` | B | --
`visdat` | A | `vis_dat`, `vis_miss`

### The `tidyverse`

The `tidyverse` package is actually a meta-package which includes the following core packages:

- `ggplot2` for creating graphics
- `dplyr` for data manipulation
- `tidyr` for creating tidy data
- `readr` for reading in rectangular data
- `purrr` for working with functions and vectors
- `tibble` for creating tibbles - lazy, surly data frames
- `stringr` for working with data strings
- `forcats` for solving problems with factors

Loading the tidyverse with `library(tidyverse)` loads those eight packages.

Installing the tidyverse also installs several other useful packages on your machine, like `glue` and `lubridate`, for example. Read more about the `tidyverse` at [https://www.tidyverse.org/](https://www.tidyverse.org/)

